Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-04-01 doc/paper.Rmd
worked on buffering paragraph

We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
This time series includes "famous" Amazonian droughts in 1997 [@mcphadenChildProdigy1997981999] and 2005[@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008] that were called "once in a century" climate events.
Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
We addressed the following questions:

@@

*Heliconia acuminata* individuals were smaller in forest fragments, although they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002], showing that probability of flowering increases with plant size and plant growth decreases with plant size <!--# other citations? -->.
Flowering probability for the largest plants is actually higher in forest fragments; however, because there are fewer large plants in fragments overall flowering probability is much lower.
The average fitted probability of flowering for plants in the upper quartile for size is 0.125 in continuous forest and 0.094 in forest fragments, equivalent to flowering once every 7.97 years and 10.63 years, respectively.
Seemingly small differences in vital rates such as these can be extremely important as they compound over many years, especially in long-lived organisms[@gagnonGrowthUnderstoryHerb2011].

@@

1.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]

    1.  We also found evidence that the reduction of microclimatic buffering by fragmentation has demographic consequences for understory plants.
    2.  Mean survival and growth rates over the entire decade were similar in continuous forest and 1-ha fragments, but were more strongly affected by variation in SPEI in fragments. As climate becomes more variable, hot, and dry, drought frequency and severity is expected to increase in the Amazon (CITE IPCC?). Our results indicate that increasing drought frequency and severity will reduce *H. acuminata* growth and survival more in fragments compared to continuous forest.
    3.  We did not see the same pattern with flowering probability, likely because flowering was overall much less likely in fragments where plants were smaller on average.

>>>>

We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
This time series includes "famous" Amazonian droughts in 1997 [@mcphadenChildProdigy1997981999] and 2005[@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008] that were called "once in a century" climate events. Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
We addressed the following questions:\
\
<
!--# Gaoue =?  Orou G. Gaoue, Carol C. Horvitz, Ulrich K. Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,https://doi.org/10.1016/j.ecolmodel.2019.03.007. -->

@@

*Heliconia acuminata* individuals were smaller in forest fragments, although they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
Size regression became more likely as plants increased in size, which is typical of long-lived forest understory plants <!--# citations -->.
Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002], showing that probability of flowering increases with plant size and plant growth decreases with plant size <!--# other citations? -->. Flowering probability for the largest plants is actually higher in forest fragments; however, because there are fewer large plants in fragments overall flowering probability is much lower. The average fitted probability of flowering for plants in the upper quartile for size is 0.125 in continuous forest and 0.094 in forest fragments, equivalent to flowering once every 7.97 years and 10.63 years, respectively. Seemingly small differences in vital rates such as these can be extremely important as they compound over many years, especially in long-lived organisms[@gagnonGrowthUnderstoryHerb2011].

@@

Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013]. It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H. acuminata* growth and survival in fragments compared to continuous forest. It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments. Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.
